\chapter{Data Preprocessing}
\begin{spacing}{1.3}

\section{Dataset Description}
We use the UNSW-NB15 network intrusion dataset from Kaggle
\cite{kaggle_unsw_nb15}. At a high level, this dataset represents network
traffic as tabular flow records rather than raw packets. Each row corresponds
to one network-flow example and includes 49 original traffic features together
with an attack category and a binary class label that indicates whether the
flow is normal or malicious. The dataset includes normal traffic as well as nine attack categories:
Fuzzers, Analysis, Backdoors, DoS, Exploits, Generic, Reconnaissance,
Shellcode, and Worms \cite{kaggle_unsw_nb15}. Since the goal of our project is
binary intrusion detection, we use the binary label and treat all attack
categories as a single positive class. The Kaggle dataset already provides predefined training and testing files, and
we use these splits directly. After duplicate removal, the training set contains 107,740 records and the test set contains 55,945 records.
Among the 49 original features, the raw dataset
contains three categorical variables, \texttt{proto}, \texttt{service}, and
\texttt{state}, together with 37 numerical traffic features. Detailed feature
descriptions can be viewed on the Kaggle dataset page
\cite{kaggle_unsw_nb15}.

\section{Data Exploration}

\subsection{Class Balance}
The training set is nearly balanced: 51,890 normal flows (48.2\%) and 55,850
attack flows (51.8\%). This means class imbalance is not a significant concern
for most models, and standard metrics such as accuracy, F1, and ROC-AUC are
informative without requiring resampling.

\subsection{Skewness and Feature Distributions}
Exploratory analysis showed that 14 numerical features were heavily right-skewed,
including \texttt{sload}, \texttt{dload}, \texttt{sbytes}, \texttt{dbytes},
\texttt{sjit}, \texttt{djit}, \texttt{rate}, \texttt{sinpkt}, \texttt{dinpkt},
\texttt{spkts}, \texttt{dpkts}, \texttt{sloss}, \texttt{dloss}, and
\texttt{response\_body\_len}. These features exhibit ranges spanning up to
$10^9$, with the vast majority of values compressed near zero. On a linear
scale, values of 1,000 and 1,000,000 appear nearly identical despite a
1,000$\times$ difference in magnitude — a difference that is highly
discriminative for detecting traffic anomalies. Figure~\ref{fig:skewness-log1p}
in Appendix~A shows the raw and log-transformed distributions for all 14 features.

\subsection{Categorical Feature Distributions}
The three categorical features contain many rare values. For example,
\texttt{proto} has 131 unique values, but only 6 have counts of at least 100.
Figure~\ref{fig:categorical-buckets} in Appendix~A shows the dominance of a
few top categories in each feature, directly motivating the bucketing approach
described in Section~2.3.

\subsection{Correlation}
Correlation analysis identified 11 feature pairs with $|r| > 0.9$, including
\texttt{dloss}--\texttt{dbytes} (0.997), \texttt{sloss}--\texttt{sbytes}
(0.996), \texttt{dpkts}--\texttt{dbytes} (0.978), \texttt{spkts}--\texttt{sbytes}
(0.966), and \texttt{tcprtt}--\texttt{synack} (0.935). No features were dropped
on this basis — tree-based models are robust to correlated inputs, and removing
features could degrade Logistic Regression's performance. The full heatmap is
shown in Figure~\ref{fig:correlation-heatmap} in Appendix~A.

\section{Data Cleaning and Feature Engineering}

\subsection{Duplicate Removal}
Both the training and test sets contained duplicate records (rows identical
across all columns except \texttt{id}). These were removed before any
preprocessing to prevent data leakage — a duplicate present in both the
training and test sets would artificially inflate evaluation metrics.

\subsection{Irrelevant Column Removal}
We remove four columns before model fitting:
\begin{itemize}
    \item \texttt{id}, because it is only a row identifier and carries no predictive signal;
    \item \texttt{attack\_cat}, because the task is binary intrusion detection rather than multi-class attack-type prediction;
    \item \texttt{stcpb} and \texttt{dtcpb}, because these TCP base sequence numbers behave more like connection identifiers than stable behavioural features.
\end{itemize}

\subsection{Categorical Bucketing and One-Hot Encoding}
Rare categorical values are bucketed into an \texttt{other} category before
one-hot encoding. We inspected the frequency distributions of
\texttt{proto}, \texttt{service}, and \texttt{state} and retained the most
common levels, summarized in
\hyperref[tab:categorical-retained]{Appendix~A, Table~\ref*{tab:categorical-retained}}.
After bucketing, one-hot encoding produces $7 + 9 + 5 = 21$ binary indicator
columns. The column set is fit on the training set and applied to the test set,
with any unseen categories filled with zeros.

\subsection{Log Transformation}
We apply $\log(1+x)$ to the 14 heavily skewed numerical features. We use
$\log(1+x)$ rather than $\log(x)$ because many features contain zero values,
and $\log(0)$ is undefined. The $\log(1+x)$ form maps zero to zero and
preserves the ordering and relative magnitudes of all values while compressing
the extreme range.

\subsection{Feature Scaling}
For models sensitive to feature scale (Logistic Regression and MLP), we apply
\texttt{StandardScaler} to all numerical features after log transformation.
The scaler is fit exclusively on the training set and applied to the test set
using the saved mean and variance, preventing any test-set information from
leaking into the scaling parameters. Tree-based models (Random Forest and
XGBoost) are scale-invariant by construction — splits depend only on feature
ordering, not magnitude — so no scaling is applied for these models.

All preprocessing artifacts (fitted scaler, one-hot column sets, column lists)
are serialized using \texttt{joblib} and reused for test set transformation,
ensuring a consistent and leakage-free pipeline.

\section{Data Representation and Final Datasets}
We maintain two model-specific representations derived from the same cleaned
data:
\begin{itemize}
    \item \textbf{\texttt{df\_model}}: 21 one-hot encoded categorical indicators and 37 scaled numerical features (log-transformed where applicable, then standardized). Used by Logistic Regression and MLP.
    \item \textbf{\texttt{df\_model\_tree}}: the same 21 one-hot encoded indicators combined with the 37 raw (unscaled) numerical features. Used by Random Forest and XGBoost.
\end{itemize}

Both matrices contain 58 predictors plus the label column.

For hyperparameter tuning, we keep the predefined Kaggle training/testing
split fixed and use \texttt{RandomizedSearchCV} with 3-fold cross-validation
on the training split only. The test split remains untouched until final
evaluation. Table~\ref{tab:data-split-summary} reports the final model-ready
dataset sizes after duplicate removal.

\begin{table}[htbp]
    \centering
    \small
    \setlength{\tabcolsep}{4pt}
    \renewcommand{\arraystretch}{0.95}
    \caption{Final model-ready dataset sizes after duplicate removal.}
    \label{tab:data-split-summary}
    \begin{tabularx}{\linewidth}{l c c c}
        \toprule
        Split & Total Rows & Normal & Attack \\
        \midrule
        Training & 107{,}740 & 51{,}890 (48.2\%) & 55{,}850 (51.8\%) \\
        Test     &  55{,}945 & 34{,}206 (61.1\%) & 21{,}739 (38.9\%) \\
        \bottomrule
    \end{tabularx}
\end{table}

\end{spacing}
